---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Loading and preprocessing the data
```{r echo=FALSE}
setwd("M:\\MyLife\\Courses\\Johns Hopkins Data Analytics\\05 Reporducable Research\\RepData_PeerAssessment1")
```

### Read in the file and tidy the data.
```{r}
steps <- read.table("activity.csv", header=TRUE, sep=",")
steps$date <- as.Date(steps$date)
```

## What is mean total number of steps taken per day?
```{r}
step.days <- aggregate(steps ~ date, steps, sum, rm.na=TRUE)
```

### The Totals by Day
```{r}
step.days
```

### Histogram of Steps Per Day
```{r}
steps.mean <- round(mean(step.days$steps),1)
steps.median <- round(median(step.days$steps),1)
steps.sd <- round(sd(step.days$steps), 1)

hist(step.days$steps, breaks=20, xlab="", ylab="", main="Histogram of Steps Per Day")
axis(1, tck=1, lty=3, col="lightgray")
axis(2, tck=1, lty=3, col="lightgray")
hist(step.days$steps, breaks=20, col="lightgreen", lwd=3, add=TRUE)
axis(1, lty=3, col="black")
axis(2, lty=3, col="black")
```

### Compute the Mean and the Median
```{r}
print(paste0("Mean = ", steps.mean, ", Median = ", steps.median))
```

## What is the average daily activity pattern?
```{r}
step.intervals <- aggregate(steps ~ interval, steps, mean, rm.na=TRUE)
step.intervals$steps <- round(step.intervals$steps, 1)
plot(step.intervals$interval, step.intervals$steps, type="n", xlab="Interval", ylab="Average Steps", main="Average Steps Per Interval")
axis(1, tck=1, lty=3, col="lightgray")
axis(2, tck=1, lty=3, col="lightgray")
lines(step.intervals$interval, step.intervals$steps, col="blue", lwd=2)
axis(1, lty=3, col="black")
axis(2, lty=3, col="black")
```

### Find the interval with the highest daily average
```{r}
max.steps <- step.intervals$steps[1]
max.interval <- step.intervals$interval[1]

for (i in 2:nrow(step.intervals)) {
    if (step.intervals$steps[i] > max.steps) {
        max.steps <- step.intervals$steps[i]
        max.interval <- step.intervals$interval[i]
    }
}

print(paste0("Max Average Steps = ", max.steps, ", At Interval = ", max.interval))
```

## Imputing missing values

### Create a new dataset "imputed.steps" from "steps" by imputing the NA values.
```{r}
imputed.steps <- steps
total.missing <- 0

for (i in 1:nrow(imputed.steps)) {
    
    # If the interval is an NA, the look up the mean for the 5 minute interval (we computed this 
    # earlier) and increment the counter keeping track of the total number of NA's.
    
    if (is.na(imputed.steps$steps[i])) {
        j <- which(step.intervals$interval == imputed.steps$interval[i])
        imputed.steps$steps[i] <- step.intervals$steps[j]
        total.missing <- total.missing + 1
    }
}
```

### Number of rows with missing steps (i.e., NA's)
```{r}
print(paste0("Total Missing Rows = ", total.missing))
```

```{r}
imputed.step.days <- aggregate(steps ~ date, imputed.steps, sum)
```

### Histogram of Imputed Steps Per Day
```{r}
hist(imputed.step.days$steps, breaks=20, xlab="Imputed Steps Per Day", 
     main="Histogram of Imputed Steps Per Day")
axis(1, tck=1, lty=3, col="lightgray")
axis(2, tck=1, lty=3, col="lightgray")
hist(imputed.step.days$steps, breaks=20, col="lightgreen", lwd=3, add=TRUE)
axis(1, lty=3, col="black")
axis(2, lty=3, col="black")
```

### Compute the Mean and the Median of Imputed Data
```{r}
imputed.steps.mean <- round(mean(imputed.step.days$steps),1)
imputed.steps.median <- round(median(imputed.step.days$steps),1)
                              
print(paste0("Mean = ", imputed.steps.mean, ", Median = ", imputed.steps.median))
```

### Differences in mean and median between the raw data (with NA's) and imputed data.
```{r}
diff.steps.mean <- round(imputed.steps.mean - steps.mean,1)
diff.steps.median <- round(imputed.steps.median - steps.median,1)

print(paste0("Difference in Mean = ", diff.steps.mean, ", Difference in Median = ", diff.steps.median))
```

## Are there differences in activity patterns between weekdays and weekends?

### Create a factor variable in the imputed data frame.

```{r}
library(lubridate)

# Set the period default to Weekday.

imputed.steps <- cbind(imputed.steps, as.factor(rep("Weekday", times=nrow(imputed.steps))))
names(imputed.steps)[4] <- "period"

levels(imputed.steps$period) <- c("Weekday", "Weekend Days")        # Setup the factor with two levels.
x <- wday(imputed.steps$date) == 6 | wday(imputed.steps$date) == 7  # Find all Saturdays and Sundays.
imputed.steps$period[which(x)] <- "Weekend Days"                    # Set the period to "Weekend Days".

```

### Panel Plot the Weekday and Weekend day  

```{r}
# This is a common plot function

plot.ss <- function(datasubset) {
    
    ss <- subset(imputed.steps, period == datasubset)                   # Subset the data.
    
    ss.intervals <- aggregate(steps ~ interval, ss, mean, rm.na=TRUE)   # Aggregate the subset by interval.
    ss.intervals$steps <- round(ss.intervals$steps, 1)
    
    # Plot the data.
    
    plot(ss.intervals$interval, ss.intervals$steps, type="n", xlab="Interval", 
        ylab="Average Steps", main=datasubset, xaxt="n", yaxt="n")
    axis(1, tck=1, lty=3, col="lightgray", cex.axis=0.75)
    axis(2, tck=1, lty=3, col="lightgray", cex.axis=0.75)
    lines(ss.intervals$interval, ss.intervals$steps, col="blue", lwd=2, xaxt="n", yaxt="n")
    axis(1, lty=1, col="black", cex.axis=0.75)
    axis(2, lty=1, col="black", cex.axis=0.75)
}

# Start a new panel plot.

plot.new()
par(mfcol=c(2,1), pin=c(5,1.5))               # Setup for two rows, one column for the two plots.
plot.ss("Weekday")              # Plot the average weekday activity by interval.
plot.ss("Weekend Days")         # Plot the average weekend day activity by interval.
```